\bf Exploring probability distributions for bivariate temporal granularities

### 2019-12-02 gravitas/gravitas.tex
aperiodic completed

---
title: Exploring probability distributions for bivariate temporal granularities
authors:
- name: Sayani Gupta
  affiliation: Department of Econometrics and Business Statistics, Monash University
  email: Sayani.Gupta@monash.edu

- name: Rob J Hyndman
  affiliation: Department of Econometrics and Business Statistics, Monash University

- name: Dianne Cook
  affiliation: Department of Econometrics and Business Statistics, Monash University

keywords:
- data visualization
- statistical distributions
- time granularities
- periodicties
- grammar of graphics
- R

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that need to be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear time granularities, circular time granularities and aperiodic calendar categorizations. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Circular time granularities accommodate periodicities in time such as hour of the day, and day of the week. Aperiodic calendar categorizations are neither linear nor circular, such as day of the month or public holidays.

@@

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2[@Wickham2009-pk].

bibliography: bibliography.bib
output: rticles::asa_article
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
library(tidyverse)
library(lubridate)
library(lvplot)
library(ggridges)
library(tsibble)
library(gravitas)
library(tsibble)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = "figure/", fig.align = "center", fig.show = "hold",
  cache = FALSE, cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
knitr::opts_knit$set(root.dir = here::here())
```

# Introduction

<!--temporal granularities and why should we care -->

<!-- problem of analyzing data with such finer scales and classical ways to deal with it -->

Temporal data are available at various resolutions depending on the context. Social and economic data like GDP is often collected and reported at coarse temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities[@aigner2011visualization].
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore all possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

<!-- # existing ways and challenges -->

Calendar-based graphics[@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs. cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in [@Bettini1998-ed]. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular or aperiodic depending on if mappings between granularities are regular or irregular. Examples of circular can be hour of the day, day of the week and that of aperiodic granularities can be day of the month or public holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

<!--motivation -->

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates [@2012-la],  of usage across households, total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data can be benefited from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. However, the motivation came through the smart meter example, this is a problem which relates to any data that needs to be analyzed for different periodicities.

<!-- at finer or coarser scales can be benefited from the approach of Exploratory Data Analysis (EDA). EDA calls for utilizing visualization and transformation to explore data systematically. It is a process of generating hypothesis, testing them and consequently refining them through investigations. -->

<!--existing closely related work and how ours relate --> 
 <!-- This paper utilizes the nestedness of time granularities to obtain multiple-order-up granularities from single-order-up ones. -->

<!--actions on multiple order up - usage -->

<!-- Finally, visualizing data across single/multiple order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to iteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.  -->

<!-- challenges and how you are dealing wih it --> 
This work provides tools for systematically exploring bivariate granularities within the tidy workflow through the following:

<!-- Pairs of granularities are categorized as either a *harmony* or *clash*, where harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. -->

  * Formal characterization of cyclic granularities

  * Facilitate manipulation of single and multiple order-up time granularities

  * Checking feasibility of creating plots or drawing inferences from any two cyclic granularities

  * Recommend prospective probability distributions for exploring distributions of univariate dependent variable across pair of granularities

  <!-- Pairs of granularities can be categorized as either a *harmony* or *clash*, where harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. -->

@@

# Linear Time granularities{#sec:timegrandef}

<!-- general calendar categorization -->
Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}.

## Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement}

The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months. However, periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, time granularities can also be **circular** or **aperiodic** depending on if the mapping of the linear granularities is regular or irregular. Examples of circular time granularities include hour of the day, and day of the week, whereas examples for aperiodic time granularities can be day of the month or public holidays.

<!-- <!-- Representation with tsibble -->
<!-- Suppose we have a tsibble with a time index in one column and keys and variables in other columns. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of row numbers (the index set) to the time index. A linear granularity is a mapping of row numbers to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. -->

<!-- <!-- What is calendar categorization --> 
<!-- What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as working day/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a **calendar categorization**.  -->

<!-- What is circular granularity and some examples -->
<!-- We specify the circular categorizations using modular arithmetic and call these **circular granularities**. The number of categories is essentially the periodicity of a circular time granularity. For example, suppose the time index is in minutes, and let $n_i$ be the number of categories created by the circular granularity $C_i$. Then the following categorizations can be computed. -->

@@

### Linear {#sec:linear-gran-def}

There has been several attempts to provide the framework for formally characterizing time-granularities. One of the first attempts occur in [@Bettini1998-ed] with the help of the following definitions:

@@

A time domain can be **discrete** (if there is unique predecessor and successor for every element except for the first and last one in the time domain), or it can be **dense** (if it is an infinite set). A time domain is assumed to be discrete for the purpose of our discussion.

@@

**Dicussion:** Each non-empty subset $G(i)$ is called a **granule**, where $i$ is one of the indexes and $G$ is a linear granularity. The first condition implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. If we consider the chronon [@aigner2011visualization] as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.

```{r linear-time,echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear granularities"}
knitr::include_graphics("Figs/linear-time.png")
```

#### Relationships between two linear granularities and association to Periodicity

[@Bettini1998-ed] talks about the relationships of linear time granularities and structure of a calendar and also relates them to the notion of periodicity in time.

@@

According to this definition, $G \trianglelefteq H$ if each granule H (j) is the union of some
granules of G. For example, $Days \trianglelefteq Months$. However, this  relationship is not fully described until we associate it with periodicity. For example, $Days \trianglelefteq Months,$, since each month is a grouping of a number of days, however it is also a periodical grouping. If leap year is ignored, the periodicity of the grouping is 1 year, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, the period becomes 400 years. 
\begin{timedom}\label{def:periodical}
A granularity H is periodical with respect to a granularity G if
(1) G \trianglelefteq H, and
(2) there exist R,P \epsilon Z+, where R is less than the number of granules of H, such
that for all i \epsilon Z, if H(i) = \bigcup_{j \in S}G(j) and H (i + R) \neq \phi then
H (i + R) = \bigcup_{j \in S} G(j + P).
\end{timedom}

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_(R-1)}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_{R-1}}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H (j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$.

This formula can be explained observing that H (j ) is either one of H (0), . . . , H (R - 1) or, for the periodicity of H, is one of these granules “shifted” ahead or behind on the time line of a finite number of periods. If $H (j ) \epsilon {H (0), . . . , H (R - 1)}$, then the formula ‘j mod R’ defines the index (among those in {0,...,R - 1}) of the granule that must be shifted to obtain H (j).The number of periods each granule of G composing H (j mod R) should be shifted is given by $\lfloor j/R \rfloor$.

Granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way [@Bettini2000-vy].

\begin{timedom}\label{def:quasi}
A granularity H is quasi-periodical with respect to a granularity G if
(1) G \trianglelefteq H, and
(2) there exist a set of intervals E1,...,Ez (the granularity exceptions) and positive
integers R,P, where R is less than the minimum of the number of granules of H
between any 2 exceptions, such that for all i \epsilon Z, if H(i) = \bigcup_{k \in [0,k]} G(j_r)  and H (i + R) \neq \phi and i + R < min(E), where E is the closest existing exception after H (i)^2, then H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P ).
\end{timedom}

@@

**Example:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

<!-- Most practical problems seem to require only a granularity system containing a set of time granularities which are all periodical (or quasi-periodical) with respect to a basic granularity. The assumption is there should be one bottom granularity and all the other granularities are generated from the bottom granularity by calendar algebraic operations. -->

#### Computation of linear time granularities

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra[@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and Calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
Thus, the relationship between the operand(s) and the resulting granularities are encoded
in the operations.

The calendar algebra consists of two kinds of operations: grouping-oriented and
granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. Example can be to consider a calendar with only two linear granularities minute and hour and hour is generated by grouping every 60 minutes. The granule-oriented operations do not change the granules of a granularity, but rather make choices of which granules should remain in the new granularity. For example, one can choose to look at the granularity "Monday" and hence select only Mondays while looking at the linear granularity "day".

@@

- **The grouping operation** : Let G be a full-integer labeled granularity, and m a positive integer. The grouping operation $Group_m(G)$ generates a new granularity G, by partitioning the granules of G into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G = Group_m(G)$ is the full-integer labeled granularity such that for
each integer i, $G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)$.

**Example** :

    - $minute = Group_{60}(second)$
    - $hour = Group_{60}(minute)$,
    - $day = Group_{24}(hour)$,
    - $week = Group_{7}(day)$,

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and l, k, m integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the lth granule of each group has |k| additional (or fewer when k < 0) granules of G2.

**Example** :

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Exploring probability distributions for bivariate temporal granularities}

  \author{
        Sayani Gupta \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Rob J Hyndman \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Dianne Cook \\
    Department of Econometrics and Business Statistics, Monash University\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Exploring probability distributions for bivariate temporal granularities}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Recent advances in technology greatly facilitates recording and storing
data at much finer temporal scales than was previously possible. As the
frequency of time-oriented data increases, the number of questions about
the observed variable that need to be addressed by visual representation
also increases. We propose some new tools to explore this type of data,
which deconstruct time in many different ways. There are several classes
of time deconstructions including linear time granularities, circular
time granularities and aperiodic calendar categorizations. Linear time
granularities respect the linear progression of time such as hours,
days, weeks and months. Circular time granularities accommodate
periodicities in time such as hour of the day, and day of the week.
Aperiodic calendar categorizations are neither linear nor circular, such
as day of the month or public holidays.

@@

Visualizing data across granularities which are either single-order-up
or multiple-order-up or periodic/aperiodic helps us to understand
periodicities, pattern and anomalies in the data. Because of the large
volume of data available, using displays of probability distributions
conditional on one or more granularities is a potentially useful
approach. This work provides tools for creating granularities and
exploring the associated time series within the tidy workflow, so that
probability distributions can be examined using the range of graphics
available in ggplot2\citep{Wickham2009-pk}.
\end{abstract}

\noindent%
{\it Keywords:} data visualization, statistical distributions, time granularities, periodicties, grammar of graphics, R
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Temporal data are available at various resolutions depending on the
context. Social and economic data like GDP is often collected and
reported at coarse temporal scales like monthly, quarterly or annually.
With recent advancement in technology, more and more data are recorded
at much finer temporal scales. Energy consumption is collected every
half an hour, while energy supply is collected every minute and web
search data might be recorded every second. As the frequency of data
increases, the number of questions about the periodicity of the observed
variable also increases. For example, data collected at an hourly scale
can be analyzed using coarser temporal scales like days, months or
quarters. This approach requires deconstructing time in various possible
ways called time granularities\citep{aigner2011visualization}. It is
important to be able to navigate through all of these temporal
granularities to have multiple perspectives on the periodicity of the
observed data. This idea aligns with the notion of EDA
\citep{Tukey1977-jx} which emphasizes the use of multiple perspectives
on data to help formulate hypotheses before proceeding to hypothesis
testing. Visualizing probability distributions conditional on one or
more granularities is a potentially useful approach for exploration.
Analysts are expected to iteratively explore all possible choices of
time granularities for comprehending possible periodicities in the data.
But too many choices and a lack of a systematic approach to do so might
become overwhelming.

Calendar-based graphics\citep{wang2018calendar} are useful in
visualizing patterns in the weekly and monthly structure well and are
capable of checking the weekends or special days. Any sub-daily
resolution temporal data can also be displayed using this type of
faceting \citep{Wickham2009-pk} with days of the week, month of the year
and another sub-daily deconstruction of time. But calendar effects are
not restricted to conventional day-of-week, month-of-year ways of
deconstructing time. There can be several classes of time
deconstructions, viz. based on the arrangement (linear vs.~cyclic) or
hierarchical order of the calendar. Linear time granularities respect
the linear progression of time such as hours, days, weeks and months.
One of the first attempts of characterizing these granularities occur in
\citep{Bettini1998-ed}. The definitions and rules defined are inadequate
to reflect periodicities in time. Hence, there is a need to define
cyclic time granularities in a different approach, which can be useful
in visualizing periodic behavior. Cyclic time granularities can be
circular or aperiodic depending on if mappings between granularities are
regular or irregular. Examples of circular can be hour of the day, day
of the week and that of aperiodic granularities can be day of the month
or public holidays. Time deconstructions can also be based on the
hierarchical structure of time. For example, hours are nested within
days, days within weeks, weeks within months, and so on. Hence, it is
possible to construct single-order-up granularities like second of the
minute or multiple-order-up granularities like second of the hour.
lubridate \citep{G_Grolemund2011-vm} creates easy access and
manipulation of common date-time objects. But most accessor functions
are limited to single-order-up granularities.

The motivation for this work comes from the desire to provide methods to
better understand large quantities of measurements on energy usage
reported by smart meters in households across Australia, and indeed many
parts of the world. Smart meters currently provide half-hourly use in
kWh for each household, from the time that they were installed, some as
early as 2012. Households are distributed geographically and have
different demographic properties such as the existence of solar panels,
central heating or air conditioning. The behavioral patterns in
households vary substantially, for example, some families use a dryer
for their clothes while others hang them on a line, and some households
might consist of night owls, while others are morning larks. It is
common to see aggregates \citep{2012-la}, of usage across households,
total kWh used each half hour by state, for example, because energy
companies need to understand maximum loads that they will have to plan
ahead to accommodate. But studying overall energy use hides the
distributions of usage at finer scales, and making it more difficult to
find solutions to improve energy efficiency. We propose that the
analysis of smart meter data can be benefited from systematically
exploring energy consumption by visualizing the probability
distributions across different deconstructions of time to find regular
patterns/anomalies. However, the motivation came through the smart meter
example, this is a problem which relates to any data that needs to be
analyzed for different periodicities.

This work provides tools for systematically exploring bivariate
granularities within the tidy workflow through the following:

\begin{itemize}
\item
  Formal characterization of cyclic granularities
\item
  Facilitate manipulation of single and multiple order-up time
  granularities
\item
  Checking feasibility of creating plots or drawing inferences from any
  two cyclic granularities
\item
  Recommend prospective probability distributions for exploring
  distributions of univariate dependent variable across pair of
  granularities
\end{itemize}

@@

\hypertarget{sec:timegrandef}{%
\section{Linear Time granularities}\label{sec:timegrandef}}

Often we partition time into months, weeks or days to relate it to data.
Such discrete abstractions of time can be thought of as time
granularities \citep{aigner2011visualization}. Examples of time
abstractions may also include day-of-week, time-of-day, week-of-year,
day-of-month, month-of-year, working day/non-working day, etc which are
useful to represent different periodicities in the data. Classes of time
abstractions may be based on arrangement or hierarchical order of time,
discussed in further details in \autoref{sec:arrangement} and
\autoref{sec:order}.

\hypertarget{sec:arrangement}{%
\subsection{Arrangement: Linear vs.~Circular
vs.~aperiodic}\label{sec:arrangement}}

The arrangement of the time domain can result in deconstructing time in
linear, circular or aperiodic (nearly circular) ways. Time granularities
are \textbf{linear} if they respect the linear progression of time.
Examples include hours, days, weeks and months. However, periodicity is
very common in all kinds of data. Time granularities which can
accomodate periodicities can be constructed by relating two linear
granularities. The mappings between linear granularities can be regular
or irregular. For example, a regular mapping exists between minutes and
hours, where 60 minutes always add up to 1 hour. On contrary, an
irregular mapping exists between days and months, since one might can
have days ranging from 28 to 31. Hence, time granularities can also be
\textbf{circular} or \textbf{aperiodic} depending on if the mapping of
the linear granularities is regular or irregular. Examples of circular
time granularities include hour of the day, and day of the week, whereas
examples for aperiodic time granularities can be day of the month or
public holidays.

@@

\hypertarget{sec:linear-gran-def}{%
\subsubsection{Linear}\label{sec:linear-gran-def}}

There has been several attempts to provide the framework for formally
characterizing time-granularities. One of the first attempts occur in
\citep{Bettini1998-ed} with the help of the following definitions:

@@

A time domain can be \textbf{discrete} (if there is unique predecessor
and successor for every element except for the first and last one in the
time domain), or it can be \textbf{dense} (if it is an infinite set). A
time domain is assumed to be discrete for the purpose of our discussion.

@@

\textbf{Dicussion:} Each non-empty subset \(G(i)\) is called a
\textbf{granule}, where \(i\) is one of the indexes and \(G\) is a
linear granularity. The first condition implies that the granules in a
linear granularity are non-overlapping and their index order is same as
time order. \autoref{fig:linear-time} shows the implication of this
condition. If we consider the chronon \citep{aigner2011visualization} as
hourly, the time domain with T hours will have \(\lfloor T/24\rfloor\)
days, \(\lfloor T/(24*7)\rfloor\) weeks and so on.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{Figs/linear-time}

}

\caption{The time domain distributed as linear granularities}\label{fig:linear-time}
\end{figure}

\hypertarget{relationships-between-two-linear-granularities-and-association-to-periodicity}{%
\paragraph{Relationships between two linear granularities and
association to
Periodicity}\label{relationships-between-two-linear-granularities-and-association-to-periodicity}}

\citep{Bettini1998-ed} talks about the relationships of linear time
granularities and structure of a calendar and also relates them to the
notion of periodicity in time.

@@

According to this definition, \(G \trianglelefteq H\) if each granule H
(j) is the union of some granules of G. For example,
\(Days \trianglelefteq Months\). However, this relationship is not fully
described until we associate it with periodicity. For example,
\(Days \trianglelefteq Months,\), since each month is a grouping of a
number of days, however it is also a periodical grouping. If leap year
is ignored, the periodicity of the grouping is 1 year, since each month
would be defined as the grouping of the same number of days (31, 28, or
30, depending on the month) every year. Considering leap years and all
their exceptions, the period becomes 400 years.
\textbackslash{}begin\{timedom\}\label{def:periodical} A granularity H
is periodical with respect to a granularity G if (1) G
\trianglelefteq H, and (2) there exist R,P \epsilon Z+, where R is less
than the number of granules of H, such that for all i \epsilon Z, if
H(i) = \bigcup\emph{\{j \in S\}G(j) and H (i + R) \neq \phi then H (i +
R) = \bigcup}\{j \in S\} G(j + P). \textbackslash{}end\{timedom\}

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, \({S_0,...,S_(R-1)}\) describing the
granules of H within one period; (ii) the value of P; (iii) the indexes
of first and last granules in H, if their value is not infinite. Then,
if \({S_0,...,S_{R-1}}\) are the sets of indexes of G describing
\(H (0), . . . , H (R - 1)\), respectively, then the description of an
arbitrary granule H (j) is given by:
\(\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)\).

This formula can be explained observing that H (j ) is either one of H
(0), . . . , H (R - 1) or, for the periodicity of H, is one of these
granules ``shifted'' ahead or behind on the time line of a finite number
of periods. If \(H (j ) \epsilon {H (0), . . . , H (R - 1)}\), then the
formula `j mod R' defines the index (among those in \{0,\ldots{},R -
1\}) of the granule that must be shifted to obtain H (j).The number of
periods each granule of G composing H (j mod R) should be shifted is
given by \(\lfloor j/R \rfloor\).

Granularities can be periodical with respect to other granularities,
except for a finite number of spans of time where they behave in an
anomalous way \citep{Bettini2000-vy}.

\textbackslash{}begin\{timedom\}\label{def:quasi} A granularity H is
quasi-periodical with respect to a granularity G if (1) G
\trianglelefteq H, and (2) there exist a set of intervals E1,\ldots{},Ez
(the granularity exceptions) and positive integers R,P, where R is less
than the minimum of the number of granules of H between any 2
exceptions, such that for all i \epsilon Z, if H(i) = \bigcup\emph{\{k
\in [0,k]\} G(j\_r) and H (i + R) \neq \phi and i + R \textless{}
min(E), where E is the closest existing exception after H (i)\^{}2, then
H (i + R) = \bigcup}\{k \in [0,k]\} G(j\_r + P ).
\textbackslash{}end\{timedom\}

@@

\textbf{Example:} Given the set of all granularities defined over the
time domain (Z; \textless{}), and the granularity relationship
\(\trianglelefteq\) (groups into), the granularity mapping each index
into the corresponding instant (same integer number as the index) is a
bottom granularity with respect to \(\trianglelefteq\).

\hypertarget{computation-of-linear-time-granularities}{%
\paragraph{Computation of linear time
granularities}\label{computation-of-linear-time-granularities}}

Linear time granularities are computed through an algebraic
representation for time granularities, which is referred to as calendar
algebra\citep{Ning2002-tf}. It is assumed that there exists a ``bottom''
granularity and Calendar algebra operations are designed to generate new
granularities from the bottom one or recursively, from those already
generated. Thus, the relationship between the operand(s) and the
resulting granularities are encoded in the operations.

The calendar algebra consists of two kinds of operations:
grouping-oriented and granule-oriented operations. The grouping-oriented
operations combine certain granules of a granularity together to form
the granules of the new granularity. Example can be to consider a
calendar with only two linear granularities minute and hour and hour is
generated by grouping every 60 minutes. The granule-oriented operations
do not change the granules of a granularity, but rather make choices of
which granules should remain in the new granularity. For example, one
can choose to look at the granularity ``Monday'' and hence select only
Mondays while looking at the linear granularity ``day''.

@@

\begin{itemize}
\tightlist
\item
  \textbf{The grouping operation} : Let G be a full-integer labeled
  granularity, and m a positive integer. The grouping operation
  \(Group_m(G)\) generates a new granularity G, by partitioning the
  granules of G into m-granule groups and making each group a granule of
  the resulting granularity. More precisely, \(G = Group_m(G)\) is the
  full-integer labeled granularity such that for each integer i,
  \(G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)\).
\end{itemize}

\textbf{Example} :

\begin{verbatim}
- $minute = Group_{60}(second)$
- $hour = Group_{60}(minute)$,
- $day = Group_{24}(hour)$,
- $week = Group_{7}(day)$,
\end{verbatim}

@@

\begin{itemize}
\tightlist
\item
  \textbf{The altering-tick operation} : Let G1, G2 be full-integer
  labeled granularities, and l, k, m integers, where G2 partitions G1,
  and \(1 \leq l \leq m\). The altering-tick operation
  \(Alter^{m}_{l,k} (G2, G1)\) generates a new full-integer labeled
  granularity by periodically expanding or shrinking granules of G1 in
  terms of granules of G2. The altering-tick operation modifies the
  granules of G1 so that the lth granule of each group has
  \textbar{}k\textbar{} additional (or fewer when k \textless{} 0)
  granules of G2.
\end{itemize}

\textbf{Example} :
